SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-19 design.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

\section{NATCH Architecture}
\label{Design}

In this section we describe our implementation of NATCH.
NATCH is a generalised prefetch throttler and can be adapted to 
any existing or upcoming prefetcher with minimal modifications.
For a practical implementation, we have used SPP as the base engine prefetcher.

@@

\begin{itemize}
\item \textbf{Original Thresholds discarded}\newline In the original
  SPP design, the internal confidence was
  compared against a preset PF\_THRESHOLD (T\textsubscript{p}) to
  decide whether to accept or reject the prefetch suggestion.
  Furthermore, the accepted suggestions were thresholded against FILL
  \_THRESHOLD (T\textsubscript{f}) to decide the fill level: L2 Cache
  vs Last Level Cache.  T\textsubscript{p} and T\textsubscript{f} were
  set to 25 and 90 respectively, on the scale of 0 to 100.  In
  NATCH, the perceptron sum is used to make these decisions.  Hence
  these two thresholds are no longer needed

%%[NOTE] Can this point be removed?
\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline
  Original SPP would not consider the slots available in the L2 MSHR
  queue.  It had an assumption that at no time can the number of
  prefetches suggested by SPP could exceed the capacity of the queue.
  While it was not proven in the paper, the internal confidence
  mechanism along with the thresholds made sure that the above
  assumption was maintained.  In NATCH since there can be no
  assumption on the degree of aggressiveness of the underlying prefetcher, 
  we explicitly check that at no point
  should the number of suggested prefetches exceed L2 MSHR queue.

\item \textbf{Enhanced Filter}\newline Original SPP introduces a
  concept of Prefetch Filter, which keeps a track of up to 1024 L2
  Cache prefetches suggestions.  For NATCH, the Filter was modified to store
  meta-data like program counter, memory access address etc.  This is
  needed to train the perceptron when the result of that particular
  prefetch gets available.  At training time, this data is used to
  retrieve the state of the prefetcher when the prediction was made.

\item \textbf{Reject Filter introduced}\newline The Prefetch Filter
  mentioned above keeps a track of the prefetches that actually
  happened in the L2 Cache.  Correspondingly, we introduced a reject
  filter.  It keeps a track of the prefetches that were suggested by
  the base SPP engine but were rejected by the perceptron.  In case it
  is detected that a future demand fetch could have used this rejected
  prefetch, the perceptron weights are updated to reflect the same

@@

\subsection{The Perceptron Layer}
\label{Design-Perceptron}
As discussed earlier, the perceptron layer was introduced to act as a
throttler to the underlying prefetchh engine.  This section explores the
details of the perceptron layer and the range of features that are
ultimately used by the perceptron for correlating the prefetches.

The neural network weights are stored in separate tables for each
feature.  The indexing of each feature is done using different number
of bits.  At most 12 bits of a feature are used to index into the
perceptron table.  Certain features require more resolution power
\textit{i.e.}, full 12 bits of indexing.  On the other hand, some
features requiring lower resolution required as low as 7 bits of
indexing.  The variable indexing was determined by studying the
features and fine-tuned empirically so as to achieve a good accuracy
vs area trade-off.  Exact details are mentioned under `Area Overheads'
in Section \ref{Method-Overheads}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15.  Each feature
gets its own dedicated table.  Our proposed design uses 9 features.
As the program execution begins, all the weights are initialized to 0.

\textbf{INFERENCING}\newline In the Champsim simulator, the
prefetching method is triggered on every L2 Cache demand access.  At
that point, the prefetcher has the option to do or not do prefetch -
in case it does, then how many cache lines to prefetch.  The suggested
prefetches can be places either in L2 Cache or in Last Level Cache.
The prefetcher gets to decide this, and is usually done based on an
internal confidence mechanism.

When the base prefetcher gets triggered, it starts suggesting candidates for prefetching. 
All these
recommended prefetches are tested in the perceptron throttler before
they can access the Prefetch Filter, as in the original SPP data-path.

The perceptron looks at the micro-architectural state, or `features',
at that instant.  Each feature indexes into the hash of up to
4096-entry table dedicated for that feature.  Feature \#1 indexes into
a particular entry in table \#1 and so on.

Once all the weights are retrieved, they are added together, much like
the traditional perceptron dot product.  On obtaining the overall sum,
it is thresholded based on the preset threshold PERC\_THRESHOLD\_LO.
Only the prefetch candidates with the perceptron sum higher that the
threshold qualify for prefetching.  The prefetches that qualify the
perceptron throttler stage go through the Prefetch Filter stage.  
The Prefetch Filter eliminates any redundant
prefetch suggestions.  In NATCH, all the meta-data that is required
to recreate the features is also stored in the Prefetch Filter.  At a
later stage, when the feedback of the current prefetched line is
available, the stored data can be used to train the perceptron.

The above generated perceptron sum is also used to decide L2 Cache vs
Last Level Cache placement of the prefetch cache line.  All prefetch
candidates qualified till this stage are thresholded against
PERC\_THRESHOLD\_HI to decide the fill level. The two thresholds:
PERC\_THRESHOLD\_LO and PERC \_THRESHOLD\_HI are empirically set.

The perceptron sum can also be seen as the sum of the individual
contribution per feature.  The value of each contribution corresponds
to the extent of confidence of the final output with that given
feature.  By summing the individual contributions, the final
perceptron sum denotes the overall confidence for that prefetch
suggestion.  In a way, by thresholding the perceptron sum on two
different thresholds, we are diving the confidence scale of the
prefetch suggestion into three bins.  The first bin corresponding to
the lowest confidence leads to prefetch candidate being rejected.  The
next bin corresponds to prefetches with a moderate confidence level.
Such prefetches are directed to fill the bigger Last Level Cache and
potentially not pollute the more scarce L2 Cache.  The highest
confidence prefetches are filled in the L2 Cache.

In addition to the Prefetch Filter mentioned above, NATCH also
maintains a `Reject Filter'.  Reject filter is a 1024-entry deep 1-way
associative table.  In case that a prefetch suggestion is rejected by
perceptron throttler, it instead gets logged into the Reject Filter.
The Reject Filter keeps track of 1024 entries that were suggested by
the Pattern Table but got rejected by the perceptron throttler.  This
can be used to train the perceptron to avoid false negatives
\textit{i.e.}, cases where prediction was to reject the prefetch but
the prefetch would have been useful.

As previously stated all the meta-data describing the program state at the
instant of prefetching, is stored in the Prefetch Filter or the Reject Filter.  
This information is useful when the perceptron needs to be updated subsequently.

\textbf{TRAINING}\newline In the prefetching environment, feedback of
a prefetch is received whenever there is an eviction or a demand
access from L2 Cache.  This acts as a trigger to update all the
base prefetcher entries and to also train the perceptron throttler.
Training of the perceptron throttler involves restoring the meta-data
that was stored in Prefetch Filter or the Reject Filter.  This is done
by indexing into the corresponding filters using the cache line
address of the block being used to train. (10-bits indexing + 6-bits
tag matching).  Once the state of the program at the time of
prefetching is available, it is used to index into the perceptron
weights table.

If a demand access block which triggers the training was tagged as a
valid prefetch in the Prefetch Filter, it means that the earlier
prefetch prediction was correct.  In that case the perceptron weights
are incremented by 1 if the predicted sum does not cross a pre-defined
threshold.  In case it is a cache block eviction that led to training
and the corresponding valid entry was found in the Prefetch Filter, it
shows that the prediction made by the perceptron was wrong.
Perceptron should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to
reflect the same.  In either case, weights are saturated at -16 or
+15.

A secondary training mechanism also kicks in during demand fetches.
Before the demand access triggers the next set of prefetches, it
checks the Reject Filter for the valid entry.  In case it is a hit,
that means that the corresponding cache line was initially suggested
by the underlying prefetch engine but rejected by the perceptron throttler.
This means that perceptron should have been more confident about that
particular prefetch.  Once such a scenario is identified, the state of
the execution at the time of prefetch is retrieved from the Reject
Table.  The retrieved data is used to index into the various weights
tables of the perceptron and the corresponding values are updated by
+1, saturating between and -16 and +15.  This update reflects
increased confidence for the prediction corresponding to that
prefetch.

This mechanism allows us to solve the classical problem of exploiting
`lost opportunity'.  In prior perceptron based implementations (and in
general, prefetching algorithms), there is usually no way of knowing
the result of not prefetch a particular line.  Our two-step NATCH
architecture allows us to overcome that issue in a simple manner.

\subsection{Features used by Perceptron}
\label{Design-Features}
Here we discuss the various features which help correlate the
prefetching decision with the program behaviour
\begin{itemize}
\item \textbf{Base Address} of the demand access which triggered the
  prefetch.  Since prefetch are triggered from the L2 demand access,
  they tend to be correlated with the triggering address.

\item \textbf{Cache Line} and \textbf{Page Address}: These two
  separate features are derived from the base address which triggered
  the prefetch, as follows: base\_addr >> LOG2\_BLOCK \_SIZE and
  base\_addr >> LOG2\_PAGE\_ SIZE respectively.  The idea behind using
  three different shifted versions of the same feature is that it
  allows us to look into a wider range of bits than with a single
  version.  It also helps give more importance to the overlapping bits
  and lesser importance to most and least significance bits.  This
  approach can also eliminate destructive interference that can be
  caused by directly folding the bits into half.

\item \textbf{Program Counter XOR Depth}: PC is for the instruction
  which triggered the prefetch chain.  Depth refers to the iteration
  count of the lookahead stages.  In general, PC is not considered as
  a good basis for doing lookahead prefetching as all the prefetches
  with depth >= 1 gets aliased onto the same PC which won't be true in
  actual demand fetch case.  This feature resolves a PC into a
  different value for each lookahead depth of prefetch speculation, giving
  a more accurate correlation in lookahead cases.  This is akin to the
  concept of Virtual Program Counters\cite{VPC} introduced by Kim
  \textit{et. al.} for indirect branch prediction.

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
    PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
  PC before the instruction that triggered the current prefetch.
  Hashing together the last three PCs tell NATCH about the path
  that led to the current demand access and helps capture and
  branching information of the current basic block.  PCs are shifted
  in the increasing order of history before being hashed together.
  This is done to avoid the resultant value of zero in case 2 or more
  PCs are the same.  Additionally, obfuscating the information as it
  gets old allows us to get a wider and an approximate look into the
  program history.

\item \textbf{Program Counter XOR Delta}: This tells us if a given PC
  favours particular value(s) of delta.  As noted earlier, while PC in
  itself does not much useful information, this hash resolves PC into
  different values based on the tendency of that PC to favour a
  certain delta.  This way, dynamic nature of different instances of
  the same memory access instruction can be captured here.

\item \textbf{Confidence}: The integer confidence on the scale of 0 to
  100 which was used in the original SPP design.  While the original
  confidence is not used directly for decision making, it can still
  contribute to the final decision made.

\item \textbf{Current Signature XOR Delta}: It can be recalled from
  the discussion of SPP in Section \ref{Background-SPP} that the new signature is
  generated using the old signature and the blobk delta.  While
  `Current Signature XOR Delta' is not the exact formula for
  generating the future signature, it gives an approximate idea of the
  path that the combination of these two values can lead to.

\item \textbf{Page Address XOR Confidence}: This feature helps look at the 
tendency of each page to be prefetch friendly or prefetch averse. It 
helps resolve a page into different entries depending on its confidence 
for prefetching, which can vary during phases of a program execution.

@@

Some of the composite features are derive from simple hashing (XOR) of
two primary features.  There is always a question of usefulness of
such composite features and the `fresh' information added.  We try to
justify the choice of each feature by quantifying the contribution
made towards predicting prefetch behaviour, in Section \ref{Method-Features}.  Finally,
as noted above, each feature indexes into its independent entry of
perceptron weights.

\subsection{Generalizing NATCH for any Prefetcher}
\label{Design-Generalizing}
The above implementation of NATCH suggests that it is highly modular and can be 
adapted to be used over any base prefecher for increased prefetch accuracy. 
In our implementation, 
only two `hooks' are required between NATCH and the baseline SPP. First  
is to make sure that all the prefetch candidates of SPP pass through 
the perceptron throttler and if qualified, the metadata for perceptron indexing
be stored. Second is needed when the feedback of a prior prefetch is available in form 
of a subsequent demand hit or cache eviction. In that case, the stored meta-data needs 
to be retrieved to update the state of the perceptrons.

In general, NATCH can be adapted to a new base prefetcher with only a few modifications.
\begin{itemize}

@@

\item \textbf{Inferencing and Storing:} All prefetch recommendations to be tested 
using the perceptron inferencing algorithm.
Perceptron output: \textit{true} and \textit{false} should be saved appropriately, along with all 
the meta-data required for perceptron indexing.

\item \textbf{Retrieving and Training:} When the feedback is available, the previously 
stored meta-data can be used to re-index into the perceptron entries and increment / decrement 
the weights.

\item \textbf{Feature Selection:} Six out of the nine features we developed use information derived 
from program execution, irrespective of the baseline SPP prefetcher. Beyond that, the feature set 
can be expanded to convey any useful information from the prefetcher to the perceptron throttler.
For example, in our implementation 'Confidence', 'Current Signature XOR Delta', and 
'Page Address XOR Confidence' are such features.

>>>>

\section{PPF Architecture}
\label{Design}

In this section we describe our implementation of PPF.  PPF is a generalised
prefetch throttler and can be adapted to most prefetchers with minimal
modifications.  For a practical implementation, we have used SPP as the base
engine prefetcher.

@@

\begin{itemize}
\item \textbf{Original Thresholds discarded}\newline In the original
  SPP design, the internal confidence was compared against a preset
  PF\_THRESHOLD (T\textsubscript{p}) to decide whether to accept or reject the
  prefetch suggestion.  Furthermore, the accepted suggestions were thresholded
  against FILL \_THRESHOLD (T\textsubscript{f}) to decide the fill level: L2
  Cache vs Last Level Cache.  T\textsubscript{p} and T\textsubscript{f} were
  set to 25 and 90 respectively, on the scale of 0 to 100.  In PPF, the
  perceptron sum is used to make these decisions.  Hence these two thresholds
  are no longer needed

%%[NOTE] Can this point be removed?
% djimenez: I think this would be necessary to reproduce our results so leave it in.

\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The original
SPP would not consider the slots available in the L2 MSHR queue.  It had an
assumption that at no time can the number of prefetches suggested by SPP
exceed the capacity of the queue.  While it was not proven in the paper, the
internal confidence mechanism along with the thresholds made sure that the
above assumption was maintained.  In PPF since there can be no assumption on
the degree of aggressiveness of the underlying prefetcher, we explicitly check
that at no point should the number of suggested prefetches exceed the L2 MSHR
queue.

\item \textbf{Enhanced Filter}\newline The original SPP introduces a concept of
the Prefetch Filter that keeps a track of up to 1024 L2 cache prefetch
suggestions.  For PPF, the Filter was modified to store metadata such as program
counter, memory access address etc.  These features are needed to train the perceptron
when the result of that particular prefetch becomes available.  At training time,
this data is used to retrieve the state of the prefetcher when the prediction
was made.

\item \textbf{Reject Filter introduced}\newline The Prefetch Filter mentioned
above keeps track of the prefetches that actually happened in the L2 Cache.
Correspondingly, we introduced a reject filter.  It keeps track of the
prefetches that were suggested by the base SPP engine but were rejected by the
perceptron.  In case it is detected that a future demand fetch could have used
this rejected prefetch, the perceptron weights are updated to reflect that
misprediction.

@@

\subsection{The Perceptron Layer}
\label{Design-Perceptron}
As discussed earlier, the perceptron layer was introduced to act as a
throttler to the underlying prefetch engine.  This section explores the
details of the perceptron layer and the range of features that are ultimately
used by the perceptron for correlating the prefetches.

The weights are stored in separate tables for each
feature.  Features are indexed using different numbers
of bits.  At most 12 bits of a feature are used to index into a
weights table.  Certain features require more resolution power
\textit{i.e.}, the full 12 bits of indexing.  On the other hand, some
features requiring lower resolution require as few as 7 bits of
indexing.  The variable indexing was determined by studying the
features and fine-tuned empirically so as to achieve a good accuracy
vs area trade-off.  Exact details are mentioned under ``Area Overheads''
in Section~\ref{Method-Overheads}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15.  Each feature
has its own dedicated table.  Our proposed design uses 9 features.
As program execution begins, all the weights are initialized to 0.

\textbf{Inference}\newline In the ChampSim simulator, the prefetching method
is triggered on every L2 Cache demand access.  At that point, the prefetcher
has the option to do or not do a prefetch -- if it does, then it has a choice
in how many cache lines to prefetch.  The suggested prefetches can be either
in the L2 Cache or the last-level cache.  The prefetcher decides this
placement, and it is usually done based on an internal confidence mechanism.

When the base prefetcher is triggered, it starts suggesting candidates for
prefetching.  All these recommended prefetches are tested in the perceptron
throttler before they can access the Prefetch Filter, as in the original SPP
datapath.

The perceptron looks at the microarchitectural state, {\em i.e.} the features,
at that instant.  Each feature is hashed to form an index into a table of up to 
4096 entries dedicated for that feature.  
Feature \#1 indexes into
a particular entry in table \#1 and so on.

Once all the weights are retrieved, they are summed.  The sum is thresholded
based on the preset threshold PERC\_THRESHOLD\_LO.  Only the prefetch
candidates with perceptron sum higher that the threshold qualify for
prefetching.  The prefetches that qualify through the perceptron throttler
stage go through the Prefetch Filter stage.  The Prefetch Filter eliminates
any redundant prefetch suggestions.  In PPF, all the metadata that is required
to recreate the features is also stored in the Prefetch Filter.  At a later
stage, when the feedback of the current prefetched line is available, the
stored data is used to train the perceptron.

The perceptron sum is also used to decide L2 Cache vs Last Level Cache
placement of the prefetch cache line.  All prefetch candidates qualified till
this stage are thresholded against PERC\_THRESHOLD\_HI to decide the fill
level. The two thresholds: PERC\_THRESHOLD\_LO and PERC \_THRESHOLD\_HI are
empirically set.

The perceptron sum can also be seen as the sum of the individual contribution
per feature.  The value of each contribution corresponds to the extent of
confidence of the final output with that given feature.  By summing the
individual contributions, the final perceptron sum denotes the overall
confidence for that prefetch suggestion.  By thresholding the perceptron sum
on two different thresholds, we are dividing the confidence scale of the
prefetch suggestion into three bins.  The first bin corresponds to the lowest
confidence leads to prefetch candidate being rejected.  The next bin
corresponds to prefetches with a moderate confidence level.  Such prefetches
are directed to fill the bigger Last Level Cache and potentially not pollute
the more scarce L2 Cache.  The highest confidence prefetches are filled in the
L2 Cache.

In addition to the Prefetch Filter mentioned above, PPF also maintains a
``Reject Filter.''  The reject filter is a 1024-entry deep direct-mapped
table.  If a prefetch suggestion is rejected by perceptron throttler, it is
logged into the Reject Filter.  The filer is used to train the perceptron to
avoid false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

% djimenez: if it were previously stated then it's redundant here? do we need
% this statement?

%As previously stated all the metadata describing the program state at the
%instant of prefetching is stored in the Prefetch Filter or the Reject Filter.
%This information is useful when the perceptron needs to be updated
%subsequently.

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 Cache.  This action triggers training of the base prefetcher as well as the
perceptron throttler.  Training the perceptron throttler involves restoring
the metadata that was stored in Prefetch Filter or the Reject Filter.  This is
done by indexing into the corresponding filters using the cache line address
of the block being used to train. (10-bits indexing + 6-bits tag matching).
Once the state of the program at the time of prefetching is available, it is
used to index into the perceptron weights table.

If a demand access block that triggers the training was tagged as a valid
prefetch in the Prefetch Filter, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 if the
predicted sum does not cross a pre-defined threshold.  If a cache block
eviction led to training and the corresponding valid entry was found in the
Prefetch Filter, then the prediction made by the perceptron was wrong.  The
perceptron should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to reflect the
misprediction. In either case, weights are saturated at -16 or +15.

A secondary training mechanism also kicks in during demand fetches.  Before
the demand access triggers the next set of prefetches, Reject Filter is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the underlying prefetch engine but rejected by the
perceptron throttler.  Thus, the perceptron should have been more confident
about that particular prefetch.  Once such a scenario is identified, the state
of the execution at the time of prefetch is retrieved from the Reject Table.
The retrieved data is used to index into the various weights tables of the
perceptron and the corresponding values are updated by +1, saturating between
and -16 and +15.  This update reflects increased confidence for the prediction
corresponding to that prefetch.

This mechanism allows us to solve the classical problem of exploiting a lost
opportunity.  In prior perceptron based implementations (and in general,
prefetching algorithms), there is usually no way of knowing the result of not
prefetching a particular line.  Our two-step PPF architecture allows us to
overcome that issue.

\subsection{Features used by Perceptron}
\label{Design-Features}
Here we discuss the various features that correlate the prefetching decision
with the program behaviour.

\begin{itemize}
\item \textbf{Base Address} of the demand access that triggered the
  prefetch.  Since prefetches are triggered from the L2 demand access,
  they tend to be correlated with the triggering address.

\item \textbf{Cache Line} and \textbf{Page Address}: These two
  separate features are derived from the base address that triggered
  the prefetch, as follows: base\_addr >> LOG2\_BLOCK \_SIZE and
  base\_addr >> LOG2\_PAGE\_ SIZE respectively.  The idea behind using
  three different shifted versions of the same feature is that it
  allows us to look into a wider range of bits than with a single
  version.  It also helps give more importance to the overlapping bits
  and lesser importance to most and least significance bits.  This
  approach can also eliminate destructive interference that can be
  caused by directly folding the bits into half.

% djimenez: this is a nice intuition and explanation.

\item \textbf{Program Counter XOR Depth}: The PC is for the instruction
  that triggered the prefetch chain.  Depth refers to the iteration count of
  the lookahead stages.  In general, the PC is not considered as a good basis
  for doing lookahead prefetching as all the prefetches with depth >= 1 are
  aliased into the same PC which would not be true in an actual demand access.
  This feature resolves a PC into a different value for each lookahead depth
  of prefetch speculation, giving a more accurate correlation in lookahead
  cases.  This is akin to the concept of Virtual Program Counters~\cite{VPC}
  introduced by Kim \textit{et. al.} for indirect branch prediction.

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
  PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
  PC before the instruction that triggered the current prefetch.
  Hashing together the last three PCs tell PPF about the path
  that led to the current demand access and helps capture and
  branching information of the current basic block.  PCs are shifted
  in the increasing order of history before being hashed together.
  This is done to avoid the resultant value of zero in case 2 or more
  PCs are the same.  Additionally, obfuscating the information as it
  gets old allows us to get a wider and an approximate look into the
  program history.

\item \textbf{Program Counter XOR Delta}: This feature tells us if a given PC
  favours particular value(s) of delta.  As noted earlier, while the PC in
  itself does not convey much useful information, this hash resolves the PC
  into different values based on the tendency of that PC to favour a certain
  delta.  Thus, the dynamic nature of different instances of the same memory
  access instruction can be captured here.

\item \textbf{Confidence}: The integer confidence on a scale of 0 to
  100 that was used in the original SPP design.  While the original confidence
  is not used directly for decision making, it can still contribute to the
  final decision made.

% djimenez: this is a little vague. can you be more precise?
\item \textbf{Current Signature XOR Delta}: Recall from
  the discussion of SPP in Section \ref{Background-SPP} that the new signature
  is generated using the old signature and the block delta.  While ``Current
  Signature XOR Delta'' is not the exact formula for generating the future
  signature, it gives an approximate idea of the path that the combination of
  these two values can lead to.

\item \textbf{Page Address XOR Confidence}: This feature scores the tendency
of each page to be prefetch friendly or prefetch averse. It helps resolve a
page into different entries depending on its confidence for prefetching, which
can vary during phases of a program execution.

@@

Some of the composite features are derived from simple hashing (XOR) of two
primary features.  There is always a question of usefulness of such composite
features and the new information added.  We justify the choice of each feature
by quantifying the contribution made towards predicting prefetch behaviour, in
Section \ref{Method-Features}.  Finally, as noted above, each feature indexes
into its independent entry of perceptron weights.

\subsection{Generalizing PPF for any Prefetcher}
\label{Design-Generalizing}
The above implementation of PPF shows that it is highly modular and can be
adapted to be used over any base prefecher for increased prefetch accuracy.
In our implementation, only two hooks are required between PPF and the
baseline SPP. The first is to make sure that all the prefetch candidates of
SPP pass through the perceptron throttler and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.

In general, PPF can be adapted to a new base prefetcher with only a few modifications.
\begin{itemize}

@@

\item \textbf{Inferencing and Storing:} All prefetch recommendations to be
tested using the perceptron inferencing algorithm.  Perceptron output:
\textit{true} and \textit{false} should be saved appropriately, along with all
the metadata required for perceptron indexing.

\item \textbf{Retrieving and Training:} When the feedback is available, the
previously stored metadata can be used to re-index into the perceptron entries
and increment / decrement the weights.

\item \textbf{Feature Selection:} Six out of the nine features we developed
use information derived from program execution, irrespective of the baseline
SPP prefetcher. Beyond that, the feature set can be expanded to convey any
useful information from the prefetcher to the perceptron throttler.  For
example, in our implementation ``Confidence,'' ``Current Signature XOR Delta,''
and ``Page Address XOR Confidence'' are such features.

### 2018-11-19 intro.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

\section{INTRODUCTION}
\label{Introduction}
Processor scaling and memory scaling have been enabled
with different philosophies in mind.  While processor scaling focused
on speed improvements, memory scaling was centred around increasing
the storage capacity.  The process technology reaching scaling
limitations led to Memory Wall~\cite{MemWall} - the increasing
performance gap between the processor speed and the memory speeds.
Various techniques have been developed to combat this widening gap.
Data prefetching is one such important technique.  Data prefetching
exploits the fact that in most applications, memory accesses have a
well-defined pattern.

An ideal prefetching scheme involves capturing this memory access
patterns to predict future prefetches in a timely manner.  Other
important decisions faced by the prefetcher are when to bring in the
prefetch block of data, which level of cache hierarchy to place the
block in and finally, what block to evict to accommodate the incoming
block.  Memory accesses patterns can be as simple as fetching the next
cache line or a complex but predictable pattern, like in pointer
chasing applications.  Predicting this often requires a trade-off
between coverage and accuracy.  
Most prefetchers achieve this by maintaining an internal confidence 
mechanism by keeping a counters to track events like cache hits / miss, 
references to an entry in the table \textit{etc}. By thresholding this 
confidence against different values, the extent of aggressiveness 
can be adjusted.

For a prefetcher to capture highly
irregular access patterns, it needs to be highly aggressive (high
coverage).  That also means the prefetcher would end up recommending a
lot of useless prefetches, leading to cache pollution (low accuracy).
This effect would get even more pronounced in multi-core scenario
where the last level cache is a shared resource~\cite{Friendly}.  Conversely, a
conservative prefetcher would be highly accurate but might not make
enough useful predictions.

To this effect we propose NATCH: NeurAl prefeTCh tHrottler.  It
is an enhancement to the existing state of the art prefetcher and
overcomes the coverage vs accuracy trade-off in a robust manner.  The
idea is to use a state-of-the-art prefetcher as the base engine.
For that we choose Signature Path Prefetcher (SPP) and 
modify the design to make it as aggressive as it can get.
This enables the base engine to capture complex memory access patterns 
and go deeper into the speculation path. This
leading to increased coverage, which comes at a cost of increased DRAM
traffic.  The prefetch suggestions
recommended by the base prefetch engine pass through the perceptron based throttler. 
Over time, perceptron layer learns to correlate the prefetch recommendations 
with various available features, leading to elimination of
useless prefetches. This leads to an overall increased accuracy of
the prefetch scheme and reduction in DRAM traffic.

\vspace{1ex}This paper talks about NATCH, its merits, analysis and
future scope for expansion.  The contributions made by this paper are:

\begin{itemize}
\item This is the first time where a truly on-line learning based
  neural network model has been used when it comes to hardware data
  prefetching.  The prior works in this area either relied on program
  semantics~\cite{Semantics} or were application
  specific~\cite{Datacenter} or could not yield a significant
  improvement over the baseline prefetcher used~\cite{BadPerc}.

\item Implementing the pereptron throttler in a robust and accurate
  practical prefetching mechanism, yielding a significant performance 
  improvement as compared to the other state of the art
  prefetchers.  Moreover, the perceptron throttler learns to adapt
  itself to shared resource constraints - leading to a further increased
  performance in multi-core and lower capacity environments.

\item Defining a methodology for determining an appropriate set of 
  features to be used regardless of the underlying prefetcher used.
  More details are explained in Section \ref{Method-Features}.

@@

On a single core configuration, NATCH increased
performance of the system by 6.84\% as compared to the 
next best prefetcher, SPP.  On a multi-core
system running a mix of memory intensive subset of the SPEC 2017
traces, NATCH saw an improvement of 11.9\% over SPP for 4-core
system and 10.67\% for 8-core system.

The rest of the paper is organized as follows.  Section \ref{Background} talks about
background work and recaptures the SPP architecture.  Section \ref{Design}
discusses the detailed implementation of the perceptron mechanism and
the features used.  Section \ref{Method} consists of the evaluation methodology
and explores the feature space for perceptron learning.  Section \ref{Results} is
the performance evaluation of our prefetcher and Section \ref{Conclusion} concludes
the paper.

>>>>

\section{Introduction}
\label{Introduction}
Processor scaling and memory scaling have been enabled with different
philosophies in mind.  While processor scaling focused on speed improvements,
memory scaling was centred around increasing the storage capacity.  Process
technology reaching scaling limitations led to the Memory Wall~\cite{MemWall}
-- the increasing performance gap between processor speed and the memory
speeds.  Various techniques have been developed to combat this wide gap. Data
prefetching is one such important technique. Data prefetching exploits the
fact that in most applications, memory accesses have a well-defined pattern.

An ideal prefetching scheme woudl capture these memory access patterns to
predict future accesses in a timely manner.  Other important decisions faced
by the prefetcher are when to bring in the prefetch block of data, which level
of the cache hierarchy to place the block in, and what block to evict to
accommodate the incoming block.  Memory accesses patterns can be as simple as
fetching the next cache line or a complex pattern such as pointer chasing.
Predicting future accesses requires a trade-off between coverage and accuracy.
Most prefetchers maintain an internal confidence mechanism by keeping counters
to track events such as cache hits / misses, references to an entry in the
table, etc. By thresholding this confidence against different values, the
degree of aggressiveness can be adjusted.

For a prefetcher to capture highly irregular access patterns, it needs to be
highly aggressive (high coverage).  Thus, the prefetcher would recommend many
useless prefetches, leading to cache pollution (low accuracy).  This effect is
even more pronounced in a multi-core scenario where the last-level cache is a
shared resource~\cite{Friendly}.  Conversely, a conservative prefetcher would
be highly accurate but might not make enough useful prefetches.

We propose Perceptron-based Prefetch Filtering (PPF).  It is an enhancement to
the existing state of the art prefetcher and robustly overcomes the coverage
vs.  accuracy trade-off.  The idea is to use a state-of-the-art prefetcher as
the base engine.  For that we choose Signature Path Prefetcher (SPP) and
modify the design to make it as aggressive as possible.  This modification
enables the base engine to capture complex memory access patterns and go
deeper into the speculation path, leading to increased coverage. Normally such
aggressive prefetching would come at the cost of increased DRAM traffic.
However, prefetch suggestions recommended by the base prefetch engine are
passed through the perceptron based throttler.  Over time, perceptron layer
learns to correlate the prefetch recommendations with various available
features, leading to elimination of useless prefetches. This filtering leads
to overall increased accuracy of the prefetch scheme and reduction in DRAM
traffic.

\vspace{1ex}This paper describes PPF, giving its merits, analysis and
future scope for expansion.  The contributions include:

\begin{itemize}

\item An on-line neural model is used for hardware data
  prefetching.  Previous work in this area either relied on program
  semantics~\cite{Semantics} or were application specific~\cite{Datacenter}.
% djimenez: removing reference to this arXiv paper. we are not obligated to cite it.
%, or
%  could not yield a significant improvement over the baseline prefetcher
%  used~\cite{BadPerc}.

\item Implementing the perceptron throttler in a robust and accurate
  practical prefetching mechanism, yielding a significant performance
  improvement as compared to the other state of the art prefetchers.
  Moreover, the perceptron throttler learns to adapt itself to shared resource
  constraints, leading to further increased performance in multi-core and
  lower capacity environments.

\item Defining a methodology for determining an appropriate set of 
  features for prediction regardless of the underlying prefetcher used.
  More details are explained in Section \ref{Method-Features}.

@@

On a single core configuration, PPF increased performance by 6.84\% as
compared to the next best prefetcher, SPP.  On a multi-core system running a
mix of memory intensive subset of the SPEC 2017 traces, PPF saw an improvement
of 11.9\% over SPP for a 4-core system and 10.67\% for an 8-core system.

The paper is organized as follows. Section \ref{Background} describes
background work and summarizes SPP.  Section \ref{Design} discusses the
detailed implementation of the perceptron mechanism and the features used.
Section \ref{Method} gives the evaluation methodology and explores the feature
space for perceptron learning.  Section \ref{Results} evaluates the
performance of the prefetcher and Section \ref{Conclusion} concludes the
paper.
